Advanced Data Structures - Succinct Bit Vector

### 2024-07-03 article/article.tex
add link to github

% \pdfoutput=1 %uncomment to ensure pdflatex processing (mandatatory e.g. to submit to arXiv)
\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository
% \nolinenumbers

@@

\author{Paul Hegenberg}{Karlsruher Institut für Technologie, Germany \and \url{https://blog.hegenberg.dev} }{urlgh@student.kit.edu}{}{}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty. Please provide at least the name of the affiliation and the country. The full address is optional. Use additional curly braces to indicate the correct name splitting when the last name consists of multiple name parts.

@@

\section{Algorithm}
In the following section we will present a datastructure
that implements \code{rank}, \code{select} and \code{access} operations.
The data structure aims to use the CPU cache as efficiently as possible, by interleaving both
vector information and metadata to fit in the same cache line of 512 bit.
Our bit vector implementation consists of \code{uint64} blocks, of which we can fit 8 inside one cache line
$L_i = (\prevsum, \vec{v}_0, \dots, \vec{v}_6)$.
To speed up the computation of rank and select we need metadata, which we store in the first
64 bits of the line.
This data structure is depicted in Figure~\ref{fig:interleaved}. 
This leaves us with a memory overhead of $12.5\%$. The used metadata $\prevsum$ is an incremental sum of previous bits
set to one. For an efficient build time of the data structure we make use of the pop count $c(\vec{v})$
instruction, which counts the number of 1 bits in a subvector $\vec{v}$, in our case an \code{uint64} number.

@@

>>>>

% \pdfoutput=1 %uncomment to ensure pdflatex processing (mandatatory e.g. to submit to arXiv)
\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository
\nolinenumbers

@@

\author{Paul Hegenberg}{Karlsruher Institut für Technologie, Germany \and \url{https://github.com/paulheg/kit_advanced_data_structures}}{urlgh@student.kit.edu}{}{}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty. Please provide at least the name of the affiliation and the country. The full address is optional. Use additional curly braces to indicate the correct name splitting when the last name consists of multiple name parts.

@@

\section{Algorithm}
In the following section we will present a data structure
that implements \code{rank}, \code{select} and \code{access} operations.
The data structure aims to use the CPU cache as efficiently as possible, by interleaving both
vector information and metadata to fit in the same cache line of 512 bit.
Our bit vector implementation consists of \code{uint64} blocks, of which we can fit 8 inside one cache line
$L_i = (\prevsum, \vec{v}_0, \dots, \vec{v}_6)$.
To speed up the computation of rank and select we need metadata, which we store in the first
64 bits of the line.
This data structure is depicted in Figure~\ref{fig:interleaved}. 
This leaves us with a memory overhead of $12.5\%$. The used metadata $\prevsum$ is an incremental sum of previous bits
set to one. For an efficient build time of the data structure we make use of the pop count $c(\vec{v})$
instruction, which counts the number of 1 bits in a subvector $\vec{v}$, in our case an \code{uint64} number.

@@

\section{Code}
The code for this implementation can be found on Github in the following repository: 
\url{https://github.com/paulheg/kit_advanced_data_structures}.
